---
title: "metadata"
author: "diana baetscher"
date: "2025-04-14"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Read in existing metadata and organize

Some samples have been extracted and have ABLG numbers, others do not and have no unique keys.


```{r load-libraries}
library(tidyverse)
library(readxl)


```

```{r read-in-data}
adult_coll <- read_xlsx("../data/POP_epigenetic_metadata/POPAdultCollections1719.xlsx")

# are vial numbers unique?
adult_coll %>%
  filter(!is.na(AGE)) 

# all of the samples with ages are from 2017 in this batch.
```
439 vials and 570 NA rows.

```{r just-samples-w-ages}
adult_coll %>%
  filter(!is.na(AGE)) %>%
  ggplot(aes(x = AGE)) +
  geom_histogram(binwidth = 1)

```

Let's assume the 2019 sample ages are in other of the other data sheets:

```{r additional-metadata}
spec_2019 <- read_xlsx("../data/POP_epigenetic_metadata/MASELKO_GOA_2019_seastorm_pop specimen.xlsx") %>%
  mutate(CRUISE = as.character(CRUISE)) %>%
  mutate(SPECIMENID = as.character(SPECIMENID))

spec_2019 %>%
  filter(!is.na(AGE)) %>% # 565 samples with AGE info
  ggplot(aes(x = AGE)) +
  geom_histogram(binwidth = 1)


```

```{r 2019-samples}
# how many samples older than 30 yrs?
spec_2019 %>%
  filter(AGE > 40)

```

38 samples with ages > 30 yrs
7 samples > 40 yrs

```{r bin-samples-based-on-age}
# set seed because of random sampling
set.seed(549)

age_selected <- spec_2019 %>%
  filter(AGE < 40) %>%
  mutate(age_grp = ifelse(AGE < 10, 1, NA)) %>%
  mutate(age_grp = ifelse(AGE > 9 & AGE < 20, 10, age_grp)) %>%
  mutate(age_grp = ifelse(AGE > 19 & AGE < 30, 20, age_grp)) %>%
  mutate(age_grp = ifelse(AGE > 29 & AGE < 40, 30, age_grp)) %>%
  group_by(age_grp) %>%
  sample_n(22, replace = F)

```

```{r combine-old-and-age-sampled-fish}
# just the oldest fish
selected_sample <- spec_2019 %>%
  filter(AGE > 39) %>% # plus the age-sampled fish
  bind_rows(age_selected)

# histogram of 96 samples
selected_sample %>%
  ggplot(aes(x = AGE)) +
  geom_histogram(binwidth = 1)
```

```{r output-info-for-selected-samples}
selected_sample %>%
  write_csv("csv_outputs/POP_selected_epigenetic_samples_20250414.csv")

```



######### READ IN THE SELECTED SAMPLES FROM CHARLES #############

```{r}
first_test <- read_csv("../data/POP_selected_DNA_sample_20250513.csv")

first_test %>%
  ggplot(aes(x = AGE)) +
  geom_histogram(binwidth = 0.5)

first_test %>%
  filter(AGE > 5 & AGE < 10)
```


### Read in DNA plate 204 - POP samples for epigenetic aging ###

```{r}
plate <- read_xlsx("../data/DNA0204_Plate_Map_POP_20250415.xlsx")

gel_info <- read_xlsx("../data/DNA0204_Plate_Map_POP_20250415.xlsx", sheet = "gel_scoring")

# I did a quick re-evaluation and there were a handful of samples that I didn't agree with
gel <- gel_info %>%
  mutate(dsb = ifelse(is.na(dsb), `DNA Quality`, dsb))
```

Ok, given that it is pretty difficult to disentangle everything that Charles did, let me first look at the metadata associated with the 96 samples that he extracted. 

Thanks to Katie's SOP about checking WGS DNA quality, I figured out the non-standard gel orientation.

```{r read-in-pop-data-from-db}
# another issue is that I don't think there were necessarily ABLG numbers attached to all of the metadata when the samples were extracted.

db <- read_xlsx("../data/POP_epigenetic_metadata/20260127_export.xlsx")
head(db)

pop_db <- db %>%
  filter(SpeciesNameOLD == "Sebastes alutus" | CommonNameOLD == "pacific ocean perch") %>%
  left_join(., spec_2019, by = c("CruiseID" = "CRUISE", "AlternateID" = "SPECIMENID"))

```



```{r join-DNA-extraction-plate-to-metadata}
plate_meta <- plate %>%
  left_join(., pop_db, by = "ABLG") %>%
  select(Well, ABLG, AlternateID, AgeYRS, SEX, ExtractedBy, ExtractionDate, ExtractionPlate, ExtractionWell, DNAngul, DNAquality, DepthM, CollectionYear, StartLatitudeDD, StartLongitudeDD) %>%
  mutate(AgeYRS = as.numeric(AgeYRS))


quality_dna <- plate_meta %>%
  left_join(., gel, by = "Well") %>%
  filter(!str_detect(dsb, "No HMW"))

```
The sex data can be identified later.


```{r look-at-the-age-distribution}
quality_dna %>%
  ggplot(aes(x = AgeYRS)) +
  geom_bar(stat = "count") +
  scale_x_continuous(breaks = c(0,5,10,15,20,25,30,35,40,45,50,55,60,65,70))



```

1) include all samples >30 (13 total)
2) include 10 samples between 0-10
3) include 10 samples between 10-20
4) include 10 samples between 20-30

```{r select-samples}
older_samples_to_send <- quality_dna %>%
  filter(AgeYRS > 30)

# select one sample from each sex and age
young_samples <- quality_dna %>%
    filter(AgeYRS < 6) %>%
    filter(ABLG %in% c("47569", "793", "47574", "47577", "697"))

# set seed because of random sampling
set.seed(549)

seqmatic_selected <- quality_dna %>%
  filter(AgeYRS < 31) %>%
  filter(AgeYRS > 5) %>% 
  filter(SEX != 3) %>%
  mutate(age_grp = ifelse(AgeYRS > 5 & AgeYRS < 13, 5, NA)) %>%
  mutate(age_grp = ifelse(AgeYRS > 12 & AgeYRS < 20, 10, age_grp)) %>%
  mutate(age_grp = ifelse(AgeYRS > 19 & AgeYRS < 31, 20, age_grp)) %>%
  group_by(age_grp) %>%
  sample_n(10, replace = F)

samples_to_send <- seqmatic_selected %>%
  bind_rows(older_samples_to_send) %>%
  bind_rows(young_samples)  %>%
  mutate(SEX = ifelse(SEX == 1, "male", SEX)) %>%
  mutate(SEX = ifelse(SEX == 2, "female", SEX))


samples_to_send %>%
  ggplot(aes(x = AgeYRS, fill = SEX)) +
  geom_bar(stat = "count") +
  scale_x_continuous(breaks = c(0,5,10,15,20,25,30,35,40,45,50,55,60,65,70))

```
How many total males/females?

```{r sex-sample-size}
# that worked well!
samples_to_send %>%
  group_by(SEX) %>%
  tally()

```



```{r}
samples_to_send %>%
  arrange(Well) %>%
  filter(Well == ExtractionWell) %>%
  ungroup() %>%
  select(ExtractionPlate, ExtractionWell, ABLG, AgeYRS, DNAngul) %>%
  write_csv("csv_outputs/POP_samples_for_seqmatic_20260420.csv")


```




